A sheaf--function dictionary for algebraic tori over local fields

### 2013-10-03 shanghai.tex
Added Cor 10.2, inserted stub into the proof of 10.1 for exactness in the top center

While we have not yet fully explored the applications of such a geometrization, we want to highlight
some of the potential benefits of Theorem~\ref{thm:application} here.

Characters on $T$ have limited
functoriality: they can be pulled back along maps to $T$, but it is less clear what to do with them
in the presence of maps from $T$ to other groups.  Sheaves, on the other hand, come
equipped with a host of functors, including pushforward. 
We hope to use this to our advantage as we work toward an adaptation of Lusztig's theory of characters sheaves to the study of characters of admissible representations of reductive $p$-adic groups that extends beyond the case of $p$-adic tori. 
\Clifton{Combine these two paragraphs? It seems they make the same point. Waldspurger also uses D-L to study depth-zero distributions. Mention this, too? Mention neither?}

DeBacker and Reeder have recently applied Deligne-Lusztig representations to part of the
local Langlands correspondence. % \cite{reeder-debacker:09a}.  
After developing Deligne-Lusztig representations, 
the next step Lusztig took in studying representations
of finite groups of Lie type was to introduce character sheaves for reductive groups over $\Fq$.
We hope that our quasicharacter sheaves 
can provide a stepping stone to a similar theory for
representations of $p$-adic groups.

@@

\begin{proposition}
Extension of scalars defines a full and faithful functor 
$B_G^* : \QCb(G) \hookrightarrow \QC(G)$,
thus identifying $\QCb(G)$ with a full subcategory of $\QC(G)$.
\end{proposition}

@@

In general, $\QCb(G)$ is an essentially
proper subcategory of $\QC(G)$. Indeed, it will follow from Theorem~\ref{thm:snake}
that the restriction of $\trFrob{G} : \QCiso{G} \to G(\Fq)^*$ to the subgroup
$\QCbiso{G}$ is an isomorphism onto the subgroup of $\chi \in G(\Fq)^*$ such
that the image of $\chi$ is bounded in $\EEx$; 
see Section~\ref{sec:snake} for more details.

@@

For general $G$, there are bounded quasicharacter sheaves which are not finite.  Indeed,
it will follow from Theorem~\ref{thm:snake} that the restriction of
$\trFrob{G} : \QCiso{G} \to G(\Fq)^*$ to the subgroup $\QCfiso{G}$ is an isomorphism
onto the subgroup of $\chi \in G(\Fq)^*$ such that the image of $\chi$ is finite in $\EEx$;
see Section~\ref{sec:snake}.

@@

  Write $W$ for the Weil group $\Weil{\Fq}$.  Extensions of
  $G^0(\bFq)$ by $A(\bFq)$, such as $A(\bFq) \to B(\bFq) \to G^0(\bFq)$,
  correspond to classes in $\Ext^1_{\ZZ[W]}(G^0(\bFq), A(\bFq))$.
  Similarly, extensions of $G(\bFq)$ by $A(\bFq)$ correspond to
  classes in $\Ext^1_{\ZZ[W]}(G(\bFq), A(\bFq))$.  The map
  $G^0(\bFq) \to G(\bFq)$ induces the map
  \[
  \Ext^1_{\ZZ[W]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[W]}(G^0(\bFq), A(\bFq))
  \]
  fitting into the long exact sequence derived from applying
  the functor $\Hom(\mbox{---}, A(\bFq))$ to $G^0(\bFq) \to G(\bFq) \to \pi_0(G)(\bFq)$:
  \[
  \Ext^1_{\ZZ[W]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[W]}(G^0(\bFq), A(\bFq)) \to \Ext^2_{\ZZ[W]}(\pi_0(G)(\bFq), A(\bFq)).
  \]
  Since $W$ has cohomological dimension $1$ \cite{brown:CohomologyGrps}*{Ex. 4.3},
  $\Ext^2_{\ZZ[W]}(\pi_0(G)(\bFq), A(\bFq))$ vanishes \cite{cartan-eilenberg:HomologicalAlgebra}*{Thm. 2.6}.

  We therefore have the existence of diagram \eqref{extension-diagram}
  at the level of $\bFq$-points.  This expresses $H(\bFq)$ as a
  disjoint union of translates of $B(\bFq)$; by transport of structure
  we may take $H$ to be a group scheme over $\bFq$.  Similarly, the
  restriction of $f$ to each component of $H$ is a morphism of
  schemes, and thus $f$ is as well.  Finally, the whole diagram
  descends to a diagram of $\Fq$-schemes since the $\bFq$-points of
  the objects come equipped with continuous $W$-actions, and the
  morphisms are $\Gamma$-equivariant.
\end{proof}

@@

We saw, in Proposition~\ref{prop:functorialG}, that the trace of Frobenius
$\trFrob{G} : \QCiso{G} \to G(\Fq)^*$ is a functorial group homomorphism.
Our main result regarding quasicharacter sheaves, below, shows that
trace of Frobenius is an isomorphism for $\QC$-schemes.

%\input{snake}
\begin{theorem}\label{thm:snake}
  If $G$ is a smooth commutative group scheme locally of finite type over
  $\Fq$ and if the abelian group $\pi_0(G)(\bFq)$
  is finitely generated then the trace of Frobenius
  \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is an isomorphism of groups.
\end{theorem}

@@

  \Clifton{Refer to the new lemma, here.}
  In the proof of Proposition~\ref{prop:restriction} we saw that $G^0$ is an algebraic group over $\Fq$. % clearly connected. 
  It now follows from Proposition~\ref{prop:connected} 
  that $\ker \trFrob{G^0} =0$ and $\coker \trFrob{G^0}=0$. 
  On the other hand, it follows from Propositions~\ref{prop:etale-iso} that $\ker \trFrob{\pi_0(G)}=0$
  and $\coker \trFrob{\pi_0(G)}=0$.

  The sequence of character groups is left-exact: by
  Proposition~\ref{prop:pullback}, it is dual to the exact sequence of
  abelian groups
  \begin{equation}\label{eq:pi0k}
  \begin{tikzcd}
    1 \rar & G^0(\Fq) \rar & G(\Fq) \rar & \pi_0(G)(\Fq) \rar & 0;
  \end{tikzcd}
  \end{equation}
  since $\Hom(\ - \ ,\EEx)$ is left-exact, the sequence of
  character groups is left exact.

  The sequence of groups of isomorphism classes of quasicharacter sheaves is right-exact:   it is exact at
  $\QCiso{G^0}$ by Proposition~\ref{prop:restriction}; and
  it is exact at $\QCiso{G}$ because the left-square
  commutes (by Proposition~\ref{prop:pullback}), the bottom sequence
  is exact at $G(\Fq)^*$ and $\coker \trFrob{\pi_0(G)} =0$
  (by Proposition~\ref{prop:etale-iso}) and   \Clifton{Awkward sentence.}
  \[
  \image(\QCiso{\pi_0(G)} \to \QCiso{G}) \subseteq \ker(\QCiso{G} \to \QCiso{G^0}),
  \]
  by the last part of Proposition~\ref{prop:pullback}. Specifically,
  suppose $\qcs{L}$ is a quasicharacter sheaf on $G$ that is trivial on
  $G^0$. Then $\trFrob{\qcs{L}}\vert_{G^0(\Fq)} = \trFrob{\qcs{L}^0} =0$,
  by Proposition~\ref{prop:pullback} again. Since the sequence of
  character groups is exact at $G(\Fq)^*$, there is
  some $\chi \in \pi_0(G)(\Fq)^*$ such that the
  pull-back of $\chi$ along $G(\Fq)\to \pi_0(G)(\Fq)$ is
  $\trFrob{\qcs{L}}$. By Proposition~\ref{prop:etale-iso},
  there is a quasicharacter sheaf $\qcs{L}_\chi$ on $\pi_0(G)$, unique up
  to isomorphism, such that $\trFrob{\qcs{L}_\chi} = \chi$. By
  Proposition~\ref{prop:pullback} again, $\qcs{L}$ is isomorphic to the
  image of $\qcs{L}_\chi$ under the monoidal functor $\QC(\pi_0(G)) \to \QC(G)$.

  It now follows immediately from the snake lemma that $\trFrob{G}$ is an
  isomorphism of groups.
\end{proof}

With Theorem~\ref{thm:snake} in hand, 
we can see why neither bounded quasicharacter sheaves 
nor finite quasicharacter sheaves are up to the task of 
geometrizing characters of $G(\Fq)$ in the generality we need.  
%
Consider the case when $G$ is the discrete \'etale group scheme $\ZZ$; 
this is indeed a $\QC$-scheme.
Let $\chi : G(\Fq) \to \EEx$ be the character determined by $\chi(1) = \ell$ 
and let $\qcs{L}$ be a quasicharacter sheaf in the isomorphism class 
corresponding to the character $\chi : G(\Fq) \to \EEx$ under Theorem~\ref{thm:snake}. 
Then $\trFrob{\qcs{L}}(1) = \ell$. Suppose $\qcs{L}$ was a bounded quasicharacter sheaf. 
Then the restriction of $\qcs{L}$ to the connected component $1\in G$ 
would be a $\ell$-adic local system on $1$ of rank~one, which would provide
an $\ell$-adic character $\rho_1 : \pi_1(1,{\bar 1}) \to \EEx$ with
$\rho_1(\Frob{\Fq}) = \ell$, necessarily continuous. 
But $\pi_1(1,{\bar 1}) = \Gal(\bFq/\Fq) \iso {\hat \ZZ}$ is compact, 
while $\rho_1(\pi_1(1,{\bar 1})) \subseteq \EEx$ is not compact (it contains $\ell^\ZZ$),
contradicting the continuity of $\rho_1$.
So the Weil sheaf $\gqcs{L}$ does not descend to a local system on $G$.
It follows that the quasicharacter sheaf $\qcs{L}$ is not a bounded quasicharacter sheaf.
%
We can play the same game with the character $\chi : G(\Fq) \to \EEx$ 
defined by $\chi(1) = 1+\ell$. In this case, the corresponding quasicharacter sheaf $\qcs{L}$ is bounded,
defined at $1$ by a pro-\'etale cover of $1$, but not finite, since $1+\ell$ is not a root of unity. 
%
These simple examples illustrate why the categories $\QCb(G)$ and $\QCf(G)$ are inadequate to geometrize characters of $G(\Fq)$.

>>>>

While we have not yet fully explored the applications of such a geometrization, we want to highlight
some of the potential benefits of Theorem~\ref{thm:application} here. 
%
Characters on $T$ have limited
functoriality: they can be pulled back along maps to $T$, but it is less clear what to do with them
in the presence of maps from $T$ to other groups.  Sheaves, on the other hand, come
equipped with a host of functors, including pushforward. 
We hope to use this to our advantage as we work toward an adaptation of Lusztig's theory of characters sheaves to the study of characters of admissible representations of reductive $p$-adic groups that extends beyond the case of $p$-adic tori. 
\Clifton{Combine these two paragraphs? It seems they make the same point. Waldspurger also uses D-L to study depth-zero distributions. Mention this, too? Mention neither?}

DeBacker and Reeder have recently applied Deligne-Lusztig representations to part of the
local Langlands correspondence.  \cite{reeder-debacker:09a}.  
After developing Deligne-Lusztig representations, 
the next step Lusztig took in studying representations
of finite groups of Lie type was to introduce character sheaves for reductive groups over $\Fq$.
We hope that our quasicharacter sheaves 
can provide a stepping stone to a similar theory for
representations of $p$-adic groups.

@@

\begin{proposition}
Extension of scalars defines a full and faithful functor 
$B_G^* : \QCb(G) \hookrightarrow \QC(G)$.
\end{proposition}

@@

In general, $\QCb(G)$ is an essentially
proper subcategory of $\QC(G)$. We will identify $\QCb(G)$ with its image under $B_G^*$
and say that a quasicharacter sheaf $\qcs{L} \in \QC(G)$ is \emph{bounded}
if it is isomorphic to some $B_G^*(\qcs{L}_0, \mu_0)$.
We will see in Corollary~\ref{cor:bounded-and-finite} that the bounded quasicharacter sheaves
are precisely those whose trace of Frobenius have bounded image.

@@

For general $G$, there are bounded quasicharacter sheaves which are not finite.  Identifying
$\QCf(G)$ with its image under $B_G^* \circ L_G$, we will see in Corollary~\ref{cor:bounded-and-finite} that
a quasicharacter sheaf is finite if and only if its trace of Frobenius has finite image.

@@

  Extensions of $G^0(\bFq)$ by $A(\bFq)$, such as $A(\bFq) \to B(\bFq) \to G^0(\bFq)$,
  correspond to classes in $\Ext^1_{\ZZ[\Weil{}]}(G^0(\bFq), A(\bFq))$.
  Similarly, extensions of $G(\bFq)$ by $A(\bFq)$ correspond to
  classes in $\Ext^1_{\ZZ[\Weil{}]}(G(\bFq), A(\bFq))$.  The map
  $G^0(\bFq) \to G(\bFq)$ induces the map
  \[
  \Ext^1_{\ZZ[\Weil{}]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Weil{}]}(G^0(\bFq), A(\bFq))
  \]
  fitting into the long exact sequence derived from applying
  the functor $\Hom(\mbox{---}, A(\bFq))$ to $G^0(\bFq) \to G(\bFq) \to \pi_0(G)(\bFq)$:
  \[
  \Ext^1_{\ZZ[\Weil{}]}(G(\bFq), A(\bFq)) \to \Ext^1_{\ZZ[\Weil{}]}(G^0(\bFq), A(\bFq)) \to \Ext^2_{\ZZ[\Weil{}]}(\pi_0(G)(\bFq), A(\bFq)).
  \]
  Since $\Weil{} \cong \ZZ$ has cohomological dimension $1$ \cite{brown:CohomologyGrps}*{Ex. 4.3},
  $\Ext^2_{\ZZ[\Weil{}]}(\pi_0(G)(\bFq), A(\bFq))$ vanishes \cite{cartan-eilenberg:HomologicalAlgebra}*{Thm. 2.6}.

  We therefore have the existence of diagram \eqref{extension-diagram}
  at the level of $\bFq$-points.  This expresses $H(\bFq)$ as a
  disjoint union of translates of $B(\bFq)$; by transport of structure
  we may take $H$ to be a group scheme over $\bFq$.  Similarly, the
  restriction of $f$ to each component of $H$ is a morphism of
  schemes, and thus $f$ is as well.  Finally, the whole diagram
  descends to a diagram of $\Fq$-schemes since the $\bFq$-points of
  the objects come equipped with continuous $\Weil{}$-actions, and the
  morphisms are $\Gamma$-equivariant.
\end{proof}

@@

We saw, in Proposition~\ref{prop:functorialG}, that the trace of Frobenius
$\trFrob{G} : \QCiso{G} \to G(\Fq)^*$ is a functorial group homomorphism.
In fact, trace of Frobenius is an isomorphism when $G$ is a $\QC$-scheme.

%\input{snake}
\begin{theorem}\label{thm:snake}
  If $G$ is a smooth, commutative group scheme, locally of finite type over
  $\Fq$, and if the abelian group $\pi_0(G)(\bFq)$
  is finitely generated, then the trace of Frobenius
  \[
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is an isomorphism of groups.
\end{theorem}

@@

  The bottom row is dual to the sequence of
  abelian groups
  \begin{equation}\label{eq:pi0k}
  \begin{tikzcd}
    1 \rar & G^0(\Fq) \rar & G(\Fq) \rar & \pi_0(G)(\Fq) \rar & 0,
  \end{tikzcd}
  \end{equation}
  which is exact by Lang's theorem on connected algebraic groups over finite fields \cite{lang:56a}.
  Since $\EEx$ is divisible, $\Hom(\ - \ ,\EEx)$ is exact and thus the dual sequence of
  character groups is also exact.

  In order to apply the snake lemma, we need to show that the top row is exact at $\QCiso{G}$ and $\QCiso{G^0}$.
  The surjectivity of $\QCiso{G} \to \QCiso{G^0}$ follows from Proposition~\ref{prop:restriction}.  Now suppose that
  $\qcs{L}$ is a quasicharacter sheaf on $G$ whose restriction to $G^0$ is trivial.... \David{Missing argument, discussed this afternoon}

  Now Lemma~\ref{lem:G0alg-grp} and Proposition~\ref{prop:connected}
  imply that $\ker \trFrob{G^0} =0$ and $\coker \trFrob{G^0}=0$. 
  Similarly, Proposition~\ref{prop:etale-iso} shows that $\ker \trFrob{\pi_0(G)}=0$
  and $\coker \trFrob{\pi_0(G)}=0$.
  We may thus apply the snake lemma and conclude that
  $\ker \trFrob{G}$ and $\coker \trFrob{G}$ both vanish.
\end{proof}

With Theorem~\ref{thm:snake} in hand, 
we can see why neither bounded quasicharacter sheaves 
nor finite quasicharacter sheaves are up to the task of 
geometrizing all characters of $G(\Fq)$ for general $\QC$-schemes $G$.

\begin{corollary} \label{cor:bounded-and-finite}
Suppose $G$ is a $\QC$-scheme and $\qcs{L}$ is a quasicharacter sheaf on $G$.  Then
\begin{enumerate}
\item $\qcs{L}$ is bounded if and only if $\trFrob{L}$ has bounded image,
\item $\qcs{L}$ is finite if and only if $\trFrob{L}$ has finite image.
\end{enumerate}
\end{corollary}
\begin{proof}
\David{These examples need to be modified to provide a proof of the corollary}
Consider the case when $G$ is the discrete \'etale group scheme $\ZZ$; 
this is indeed a $\QC$-scheme.
Let $\chi : G(\Fq) \to \EEx$ be the character determined by $\chi(1) = \ell$ 
and let $\qcs{L}$ be a quasicharacter sheaf in the isomorphism class 
corresponding to the character $\chi : G(\Fq) \to \EEx$ under Theorem~\ref{thm:snake}. 
Then $\trFrob{\qcs{L}}(1) = \ell$. Suppose $\qcs{L}$ was a bounded quasicharacter sheaf. 
Then the restriction of $\qcs{L}$ to the connected component $1\in G$ 
would be a $\ell$-adic local system on $1$ of rank~one, which would provide
an $\ell$-adic character $\rho_1 : \pi_1(1,{\bar 1}) \to \EEx$ with
$\rho_1(\Frob{\Fq}) = \ell$, necessarily continuous. 
But $\pi_1(1,{\bar 1}) = \Gal(\bFq/\Fq) \iso {\hat \ZZ}$ is compact, 
while $\rho_1(\pi_1(1,{\bar 1})) \subseteq \EEx$ is not compact (it contains $\ell^\ZZ$),
contradicting the continuity of $\rho_1$.
So the Weil sheaf $\gqcs{L}$ does not descend to a local system on $G$.
It follows that the quasicharacter sheaf $\qcs{L}$ is not a bounded quasicharacter sheaf.
%
We can play the same game with the character $\chi : G(\Fq) \to \EEx$ 
defined by $\chi(1) = 1+\ell$. In this case, the corresponding quasicharacter sheaf $\qcs{L}$ is bounded,
defined at $1$ by a pro-\'etale cover of $1$, but not finite, since $1+\ell$ is not a root of unity. 
\end{proof}
